---
title: "Google Data Analytics Certification — R Practice Notebook"
subtitle: "Consolidated revision notes"
author: "Subodh"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

These are my consolidated practice notes from the **Google Data Analytics Certification**,
covering base R, the `tidyverse`, data cleaning, summarizing, and `ggplot2` visualization.
You may refer this for own learning or revision if you have already taken the course.

---

## 1. Environment Setup


```{r install-packages, eval=FALSE}
install.packages(c(
  "tidyverse", "palmerpenguins", "here", "skimr", "janitor",
  "Tmisc", "datasauRus", "SimDesign"
))
```

Packages actually loaded when the notebook is knitted:

```{r load-libraries}
library(tidyverse)     # loads ggplot2, dplyr, tidyr, readr, purrr, tibble, stringr, forcats
library(palmerpenguins) # sample dataset: penguins
library(skimr)          # quick, rich data summaries
library(janitor)        # clean_names() and other cleaning helpers
```

---

## 2. Working with Dates and Times (`lubridate`)

```{r dates-times}
today()
now()

mdy("January 20, 2023")
dmy("20-Jan-2021")
ymd_hms("2021-01-20 20:11:59")
mdy_hm("01/20/2021 08:01")
as_date(now())
```

---

## 3. Base R Fundamentals

### 3.1 Arithmetic

```{r arithmetic}
quarter_1_sales <- 3457.98
quarter_2_sales <- 1243.23

midyear_sales <- quarter_1_sales + quarter_2_sales
year_end_sales <- midyear_sales * 2

midyear_sales
year_end_sales
```

### 3.2 Conditional Logic (`if` / `else if` / `else`)

```{r conditionals}
x <- -7
if (x > 0) {
  print("x is a positive number")
} else if (x == 0) {
  print("x is zero")
} else {
  print("x is either negative or zero")
}

# Applying conditionals to real data (airquality, base R dataset)
data("airquality")

if (airquality$Temp[1] < 80) {
  print("It's not a hot day!")
} else {
  print("It's a hot day.")
}

ozone_level <- airquality[1, "Ozone"]
if (is.na(ozone_level)) {
  print("Ozone reading is missing for the first day.")
} else if (ozone_level < 30) {
  print("Low ozone level.")
} else if (ozone_level < 100) {
  print("Moderate ozone level.")
} else {
  print("High ozone level.")
}
```

### 3.3 Subsetting with Logical Operators

```{r subsetting}
high_solar_wind <- subset(airquality, Solar.R > 150 & Wind > 10)
low_solar_late  <- subset(airquality, Solar.R < 150 & Day > 20)

# Negation with '!' - everything EXCEPT rows matching the condition
not_high_solar_late <- subset(airquality, !(Solar.R > 150 & Day > 20))

head(high_solar_wind)
head(low_solar_late)
head(not_high_solar_late)
```

---

## 4. Importing & Inspecting Data

```{r inspect-data}
# Palmer Penguins
summary(penguins)
glimpse(penguins)

# diamonds (ggplot2 built-in dataset)
head(diamonds)
str(diamonds)
colnames(diamonds)

# mtcars
data(mtcars)
head(mtcars)
```

---

## 5. Data Cleaning & Wrangling

### 5.1 Renaming Columns

```{r rename}
diamonds %>% rename(carat_new = carat) %>% head()
diamonds %>% rename(carat_new = carat, cut_new = cut) %>% head()
```

### 5.2 Quick Summary Statistics

```{r summarize}
diamonds %>% summarise(mean_carat = mean(carat))
```

### 5.3 Standardizing Column Names

```{r clean-names}
penguins %>% clean_names() %>% colnames()
```

### 5.4 Adding & Removing Columns (`mutate` / `select`)

```{r mutate-select}
penguins_km <- penguins %>%
  mutate(body_mass_kg = body_mass_g / 1000,
         flipper_length_m = flipper_length_mm / 1000)

skim_without_charts(penguins_km)

# dropping a column
penguins_km %>% select(-flipper_length_m) %>% colnames()
```

### 5.5 Splitting & Combining Columns (`separate` / `unite`)

```{r separate-unite}
employee <- data.frame(
  id = 1:10,
  name = c("John Mendes", "Rob Stewart", "Rachel Abrahamson", "Christy Hickman",
           "Johnson Harper", "Candace Miller", "Carlson Landy", "Pansy Jordan",
           "Darius Berry", "Claudia Garcia"),
  job_title = c("Professional", "Programmer", "Management", "Clerical", "Developer",
                "Programmer", "Management", "Clerical", "Developer", "Programmer")
)

employee_split <- employee %>% separate(name, into = c("first_name", "last_name"), sep = " ")
employee_split %>% unite("full_name", first_name, last_name, sep = "-")
```

---

## 6. Grouping & Summarizing

```{r grouping}
# single grouping
penguins %>%
  group_by(island) %>%
  drop_na() %>%
  summarise(mean_bill_length_mm = mean(bill_length_mm))

# multiple grouping
penguins %>%
  group_by(species, island) %>%
  drop_na() %>%
  summarise(
    max_bill_length_mm = max(bill_length_mm),
    min_bill_depth_mm  = min(bill_depth_mm),
    mean_bill_length_mm = mean(bill_length_mm),
    .groups = "drop"
  )

# filtering
penguins %>% filter(species == "Adelie") %>% head()
```

### Piping (`%>%`) with `ToothGrowth`

```{r piping}
data("ToothGrowth")

# Nested (harder to read)
arrange(filter(ToothGrowth, dose == 0.5), len) %>% head()

# Piped (easier to read) — Cmd/Ctrl+Shift+M inserts %>%
ToothGrowth %>%
  filter(dose == 0.5) %>%
  arrange(len) %>%
  head()

# Grouped summary via pipe
ToothGrowth %>%
  filter(dose == 0.5) %>%
  group_by(supp) %>%
  summarise(mean_len = mean(len, na.rm = TRUE), .groups = "drop")
```

---

## 7. Checking for Data Bias (`SimDesign::bias()`)

`bias()` compares actual vs. predicted values. A result close to **zero** means the
predictions are unbiased; **positive** means the model under-predicts on average;
**negative** means it over-predicts.

```{r bias-check}
library(SimDesign)

actual_temp    <- c(68.3, 70, 72.4, 71, 67, 70)
predicted_temp <- c(67.9, 69, 71.5, 70, 67, 69)
bias(actual_temp, predicted_temp)

actual_sales    <- c(150, 203, 137, 247, 116, 287)
predicted_sales <- c(200, 300, 150, 250, 150, 300)
bias(actual_sales, predicted_sales)
```

---

## 8. Data Visualization with `ggplot2`

The "grammar of graphics" workflow:

1. `ggplot()` + point to a dataset
2. Add a geometry layer (`geom_point()`, `geom_boxplot()`, etc.)
3. Map variables inside `aes()`
4. Add labels/themes as needed
5. Use `facet_wrap()` / `facet_grid()` to split into small multiples

### 8.1 Basic Scatter Plot

```{r viz-basic}
ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```

### 8.2 Grouping by Category (color, shape, size, transparency)

```{r viz-grouping}
ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(aes(shape = species), size = 3, alpha = 0.7)
```

### 8.3 Trend Lines (Per Group vs. Overall)

```{r viz-trendlines}
# One trend line per species (color mapped globally)
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)

# Single overall trend line (color mapped only inside geom_point)
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species), size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed")
```

### 8.4 Faceting (Small Multiples)

```{r viz-facets}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species), alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, color = "gray40", linetype = "dotted") +
  facet_wrap(~species) +
  theme_minimal()

# Two-way faceting with facet_grid (NA rows removed first — ggplot can't drop them itself)
ggplot(data = na.omit(penguins)) +
  geom_point(aes(x = flipper_length_mm, y = body_mass_g, color = sex)) +
  facet_grid(species ~ sex)
```

### 8.5 Polished, Presentation-Ready Plot

```{r viz-polished}
ggplot(data = penguins,
       aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~species) +
  labs(
    title = "Relationship Between Flipper Length and Body Mass",
    subtitle = "Palmer Penguins Dataset",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    color = "Species"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5),
    legend.position = "bottom"
  )
```

### 8.6 Comparing Distributions (Boxplot, Violin, Density)

```{r viz-distributions}
ggplot(penguins, aes(x = species, y = body_mass_g, fill = species)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.15, alpha = 0.7) +
  theme_minimal() +
  labs(title = "Body Mass Distribution by Species")

ggplot(penguins, aes(x = body_mass_g, fill = species)) +
  geom_density(alpha = 0.3) +
  theme_minimal() +
  labs(title = "Density Distribution of Body Mass")
```

### 8.7 Bar Charts

```{r viz-bar}
ggplot(data = diamonds) +
  geom_bar(mapping = aes(x = color, fill = cut)) +
  facet_wrap(~cut)

# Rotated axis labels for long category names
ggplot(data = penguins) +
  geom_bar(mapping = aes(x = species, fill = species)) +
  facet_wrap(~island) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### 8.8 Annotations

```{r viz-annotate}
ggplot(data = penguins) +
  geom_point(mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  labs(
    title = "Palmer Penguins: Body Mass vs. Flipper Length",
    subtitle = "3 Species",
    caption = "Data: palmerpenguins package"
  ) +
  annotate("text", x = 220, y = 3500, label = "The Gentoos are the largest",
           color = "purple", fontface = "bold", size = 3, angle = 22)
```

### 8.9 Why Visualize? Anscombe's Quartet & the Datasaurus Dozen

These two datasets are the classic teaching example for *why summary statistics alone
are not enough* — very different-looking data can share nearly identical means,
standard deviations, and correlations.

```{r viz-quartet, warning=FALSE}
library(Tmisc)
data("quartet")

quartet %>%
  group_by(set) %>%
  summarise(mean_x = mean(x), sd_x = sd(x), mean_y = mean(y), sd_y = sd(y), cor_xy = cor(x, y))

ggplot(quartet, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~set)
```

```{r viz-datasaurus, warning=FALSE}
library(datasauRus)
data(datasaurus_dozen)

ggplot(datasaurus_dozen, aes(x = x, y = y, colour = dataset)) +
  geom_point() +
  theme_void() +
  theme(legend.position = "none") +
  facet_wrap(~dataset, ncol = 4)
```

### 8.10 Saving a Plot

```{r viz-save, eval=FALSE}
ggsave("penguin_species_plot.png")
```

---

## Summary

This notebook covers, in order: dates/times, base R conditionals and subsetting,
importing and inspecting data, `dplyr` cleaning verbs, grouping/summarizing, bias
checking, and a full `ggplot2` visualization progression from a bare scatter plot to
faceted, annotated, presentation-ready charts.
